A Container-Based Cloud-Native Architecture for the Reproducible Execution of Multi-population Optimization Algorithms

### 2020-10-16 draft.tex
More changes

Besides being, all of them, population-based, they also share the
common characteristic of creating an initial set of random candidate
solutions which is later used by a nature-inspired metaheuristic to
generate a new set of candidates; there are parts of this process that
can be easily done in parallel. For instance, the fitness of each
individual can be (in general) evaluated independently of others, and
similarly, each population could evolve in isolation. Researchers use
the term multi-population based methods when referring to techniques
using many populations as part of their optimization strategy.  Since
earlier works, researchers have been proposing some form of
parallelization \cite{muhlenbein1988evolution} to increase these
algorithms' scalability by adding some form of interaction between
populations running in isolation.  The island model was one of the
first techniques proposed for parallelization, which led to early
proofs of speed-up \cite{gorges1990explicit,grosso1985computer}.
Since then, other population-based algorithms have adopted the idea.
Researchers have found additional benefits besides the execution
speed; these include avoiding a premature convergence and maintaining
the diversity of the global population \cite{li2015multi}.

@@

Cloud computing has become the standard way of running
enterprise applications.  Not only because of the convenience of the
pay-as-you-go model or the non-existent sunk cost of physical facilities, but also because
it offers a way of describing the infrastructure as part of the code, so that it
is much easier to reproduce results and this has been a boon for scientific
computing.
However,  cloud computing has also been evolving, going from simply
putting old-style monolithic applications on virtual machines executing on an external 
data center to microservices \cite{microservices} and serverless architectures \cite{varghese2018next,Varghese2018849} 
that favor the parallel and asynchronous communication of
heterogeneous resources. In the process, a new methodology for designing 
{\em cloud-native applications} has been created, with brand new methods and
techniques \cite{Baldini2016287}. 
In these architectures, services, or even functions are seen as independent 
processing nodes, departing from the monolithic
and even distributed paradigms, and becoming a loosely coupled collection of {\em
stateless functions}\cite{malawski2017serverless}, that react to events. 
%and only come into existence while they are doing some processing.
Systems developed with the patterns outlined above have the
properties of a reactive system \cite{boner2014reactive}, and they are generally 
more flexible, fault-tolerant, and scalable. % Some properties and reference - Mario  
% Again, you have to justify the need for a reactive architecture and define it before this paragraph - JJ

@@

    \item They need to establish a workflow of local development and
prototyping while having the option of deployment to services provided
by a cloud provider, thus using modern methodologies and technologies.

@@

We organized the paper as follows: First, in Section \ref{multi}, we present a
background of the fundamental issues of integrating nature-inspired optimization
and multi-population methods. Section \ref{soa} presents state of the art relevant to
our work. In Section \ref{method}, we present the proposed method and the container-based
application in Section \ref{docker} and the workflow for reproducible experiment design in Section \ref{experiment_flow}. % No reference to section 6.
Section \ref{setup} describes the empirical evaluation design to assess the effectiveness
of the method, concerning scalability and % This is a subsection, fix it. 
the heterogeneous populations' performance.  % No references to sections 9 and 10
Finally, we offer the conclusions of this paper and suggestions for future work in Section \ref{conclusions}.

@@

In this section, we present the design
and a container-based implementation of the model we propose to execute
multi-population based algorithms. We follow the requirements outlined in
Section \ref{multi} and the general design principles of cloud-native
applications we have mentioned earlier; we intend to steer
clear of a direct translation of ``classical'' GA architectures to the
cloud and design from the ground up a cloud-native method. As a first step, we describe the general
architecture and then every component in detail.

The most basic data structure used throughout the system is the population, and at a higher
level, we can view the system as a continuous stream of populations
flowing from one process to the other. By ``continuous stream,'' we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a population to another, it must pack the population into
a message and then send the message 
to the queuing system without waiting for a response; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that will be triggered when a
message is pushed to it. At each end of the queue, the two components are 
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture and makes
it highly scalable; one reason for this is the use of functions
with no secondary effects, that is,
stateless functions. These functions do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, there is no difference
between having one or many copies of the same function pulling work from the queue 
simultaneously, because there are no side effects, including unwanted resource
locking problems resulting from this concurrency.
% This is important and we should probably mention that in the
% introduction - JJ 
Based on these general principles, we next define the proposed model.

@@

\subsection{Event-Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of producers and consumers of messages. First, we can see two queues, 
one labeled \texttt{Input} and the other one \texttt{Output}. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the queue box, and pull or pop operations as solid arrows leaving from the right side.
The architecture has at least four processes indicated in the diagram as
swimlanes: First, the \texttt{Setup} process, responsible for reading a configuration
file and creating the initial populations. Second, the \texttt{Controller} process,
responsible for the migration between populations and keeping track of the algorithm's
iterations. The \texttt{Message Queue} process runs the \texttt{Input} and \texttt{Output}
queues. Finally, there is at least one \texttt{Stateless Function} process responsible
for running the isolated algorithms. In the example, two processes \texttt{PSO} and \texttt{GA},
are shown.

@@

\item Once they finish the execution, the population state
is packed again with additional metadata about the algorithm's execution.
The resulting populations are now pushed to the Output queue. Once finished, another 
population is pulled from the queue.

\item The \texttt{Controller} process is responsible for keeping track of the progress of 
the search. It pulls current populations from the \texttt{Output} queue, inspects the metadata, 
and if an optimal solution has been found or the maximum number of iteration has been
reached, it signals the stop of the execution.

@@

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of the proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm
each island is executed in a separate processing node, i.e., in a virtual machine, CPU,
or thread. In this case, we can have just one processing node or many nodes, each 
running with a different search strategy or parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamically change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantage is that algorithm designers have many options for implementing this simple architecture. 
The same components can be implemented as a single multi-threaded program or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels for multi-threaded
execution.
\end{itemize}

@@

On the other hand, there are several caveats: It is more
costly to move entire populations as messages than only passing
specific individuals from a process to another. Designers must consider
this cost when working with large individuals, and if possible, send
populations using several messages, use compression, or adjust the size
of populations. Pool-based algorithms also suffer from this drawback,
but web-based implementations have been working with continuous
optimization problems with 1000 dimensions.  However, in other cases, % Reference that paper - Mario
this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four other
proposals for multi-population algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
populations, processing nodes, and algorithms.  In Figure \ref{fig:population_message}, we show the
main components of a message-driven architecture. We can see that the algorithm
and populations are separated, and the \texttt{Population queue} keeps the state
of populations. While algorithms only need the populations as parameters
for their execution.

@@

In contrast, in the classic island model shown in Figure
\ref{fig:classicisland}, algorithms are  population methods, and the same processing
node keeps both the algorithm process and the state of the population.  Other
processing nodes follow the same configuration and form of execution and they
only pass specific individuals between them.  In this model, adding additional
processing nodes, on-runtime, can be more difficult because they need to have
an addressing mechanism or simply be aware of these new nodes. This is
solved in part in Peer to Peer (P2P) evolutionary algorithms
\cite{juanlu:europar}, but coupling still exists, meaning that every
node must run the whole evolutionary algorithm and hold full copies of
its subpopulation.

@@

\begin{enumerate}
\item The controller continually pulls new messages from the \texttt{Output} 
message queue. These messages are instantly emitted by the  \texttt{consumed\_messages}
\texttt{Observable}. In this particular period, we can see that the stream receives 
two populations from another experiment and are shown as two red marbles.  
The filter operator removes all messages that belong to a different experiment; 
in this case, we are only interested in blue marbles.

@@

Developers can add new algorithms as stateless functions in
several ways and with two levels of integration. In the first level of
integration, the algorithm execution is not dependent on an initial
configuration created in the setup, and it only needs to take messages from the
input queue and return the modified population to the output queue. At the
second level, the algorithm receives a configuration from the initial setup,
specifying the initial algorithm parameters. We show an example of the configuration
file received by workers in Listing \ref{code:config}  in which we abbreviated
the population key (the ellipsis is used instead). The message includes the 
current state of the population and data about the problem, algorithm, and parameters.

@@

  \item[A container orchestration service]: These services are compatible with {\tt docker-compose}
    files for configuration and provide transparent provisioning. Examples of these
    services are Amazon Elastic Container Service,  Red Hat Openshift
    (which is also open source), and Marantis Docker Enterprise using Docker Swarm.
    Salza et al. \cite{salza2019speed} tested a variant of this option by using a CoreOs cluster.

@@

We chose the second option, a container orchestration service. We
deployed the EvoSwarm application in Amazon Elastic Container Service (ECS)
since it needs fewer changes in configuration (from deploying it locally using
{\tt docker-compose}) and it has a lower cost than the Kubernetes
alternative. This does not imply any loss of generality; first,
because it's {\tt docker-compose} compatible, and it can be deployed
in approximately the same way to any of the services, mentioned above,
compatible with it. Since {\tt docker-compose} is open source, it can
be ported to any computing node, and can thus be deployed to any other
cloud service like Azure or Google Cloud Platform. At any rate, please
note that there are several levels of infrastructure programming
here. At the lowest level, we are using docker containers, which are
compatible across a myriad of cloud providers in one way or other. At
the next level, we are using a container composition technology that
is also standard, {\tt docker compose}, once again compatible across
any platform that can host that kind of service. There's, however, a
final, platform-specific, infrastructure definition that will have to
vary across any public or private cloud platform that we are using,
and that will tap the cloud provider API. These changes, however, are
usually minimal, and can also be done through the CLI or using the web
console.

For instance, we can define an ECS cluster directly on the AWS Elastic Cloud VMs or through the
AWS Fargate service. Fargate is a serverless compute engine for containers that
provides automatic provision and management of AWS ECS resources. With this
service, calculating the cost of execution is simplified because users only
specify the computing units and memory for the application, instead of every
component. However, AWS Fargate is not suited for an experimental setting because
many of the details are out of user control. For this reason, we chose the
alternative of using a cluster of EC2 containers. The drawback is that we need to manually
provision the  EC2 containers, networking, and group permissions for each
experimental setup; for instance, when going from 2 to 8 workers, we need to increase the
number of EC2 instances to meet the required amount of CPU units.

>>>>

Besides being, all of them, population-based, they also share the common characteristic of
creating an initial set of random candidate solutions later used by a 
nature-inspired heuristic to generate a new set of candidates. 
Population-based algorithms are intrinsically parallel.
The fitness of each individual can be (in general) evaluated independently of 
others, and similarly, each population could evolve in isolation. Researchers use the term
multi-population based methods when referring to techniques using
many populations as part of their optimization strategy. 
Since earlier works, researchers have been proposing some form of parallelization
\cite{muhlenbein1988evolution} to increase these algorithms' scalability
by adding some form of interaction between populations running in isolation.
The island model was one of the first techniques proposed for parallelization,
which led to early proofs of speed-up \cite{gorges1990explicit,grosso1985computer}. 
% Repetitive - M
%The concept was to divide a large population into communicated subpopulations. 
Since then, other population-based algorithms have adopted the idea. 
Researchers have found additional benefits
besides the execution speed; these include avoiding a premature convergence and
maintaining the diversity of the global population \cite{li2015multi}.

@@

%% JJ Can you improve the redability of the paragraph below? 
%% Grammarly suggests that is hard to read, verly long sentences.
%% Edited a little - Mario
Cloud computing has become the standard way of running
enterprise applications.  Not only because of the convenience of the
pay-as-you-go model or the non-existent sunk cost of physical facilities, but also because
it offers a way of describing the infrastructure as part of the code, so that it
is much easier to reproduce results and this has been a boon for scientific
computing.  
However,  cloud computing has also been evolving, going from simply
putting old-style monolithic applications on virtual machines executing on an external 
data center to microservices \cite{microservices} and serverless architectures \cite{varghese2018next,Varghese2018849} 
that favor the parallel and asynchronous communication of
heterogeneous resources. In the process, a new methodology for designing 
{\em cloud-native applications} has been created, with brand new methods and
techniques \cite{Baldini2016287}. 
In these architectures, services are seen as independent processing nodes, 
departing from monolithic or distributed paradigms to becoming a loosely 
coupled collection of {\em
stateless functions}\cite{malawski2017serverless}, that react to events. 
%and only come into existence while they are doing some processing.
Systems developed with the patterns outlined above have the
properties of a reactive system \cite{boner2014reactive}, and they are generally 
more flexible, fault-tolerant, and scalable. % Some properties and reference - Mario  
% Again, you have to justify the need for a reactive architecture and define it before this paragraph - JJ

@@

    \item They need to establish a workflow of local development and
prototyping while having the option of deployment to cloud services,
 thus using modern methodologies and technologies.

@@

We organized the paper as follows: First, in Section \ref{multi}, we present a
background of the fundamental issues of integrating nature-inspired optimization
and multi-population methods. Section \ref{soa} presents state of the art relevant to
our work. In Section \ref{method}, we present the proposed method and the container-based
application in Section \ref{docker} and the workflow for reproducible experiment design in Section \ref{experiment_flow}. % No reference to section 6.
Section \ref{setup} describes the empirical evaluation design to assess the 
method's effectiveness, concerning scalability and % This is a subsection, fix it. 
the heterogeneous populations' performance.  % No references to sections 9 and 10
Finally, we offer the conclusions of this paper and suggestions for future work in 
Section \ref{conclusions}.

@@

In this section, we present the design
and a container-based implementation of the model we propose to execute
multi-population-based algorithms. We follow the requirements outlined in
Section \ref{multi} and the general design principles of cloud-native
applications we have mentioned earlier; we intend to steer
clear of a direct translation of ``classical'' GA architectures to the
cloud and design from the ground up a cloud-native method. As a first step, we describe the general
architecture and then every component in detail.

The most basic data structure used throughout the system is the population, and at a higher
level, we can view the system as a continuous stream of populations
flowing from one process to the other. By ``continuous stream,'' we mean that ideally,
everything must happen asynchronously without components needing to wait for
others. In this kind of system, this streaming functionality is normally
implemented by using a message queue system. For instance, if one process needs
to communicate a population to another, it must pack the population into
a message and then send the message 
to the queuing system without waiting for a response; this
means that after the process pushes the message, it continues its execution
without waiting for a result. On the other end, the recipient subscribes to the
message queue by defining an event handler method that will be triggered when a
message is pushed to it. At each end of the queue, the two components are 
Producers that push messages to the queue and Consumers that pull them. This
pattern is an essential component of a reactive architecture and makes
it highly scalable; one reason for this is the use of functions
with no secondary effects, that is,
stateless functions. These functions do not need to read, keep, or modify data outside of the method. If
Consumers are implemented using stateless functions, there is no difference
between having one or many copies of the same function simultaneously pulling work
from the queue. There are no side effects, including unwanted resource
locking problems resulting from this concurrency.
% This is important and we should probably mention that in the
% introduction - JJ 
Based on these general principles, we next define the proposed model.

@@

\subsection{Event-Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in our previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of producers and consumers of messages. First, we can see two queues, 
one labeled \texttt{Input} and the other one \texttt{Output}. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the queue box, and pull or pop operations as solid arrows leaving from the right side.
The architecture has at least four processes indicated in the diagram as
swimlanes: First, the \texttt{Setup} process, responsible for reading a configuration
file and creating the initial populations. Second, the \texttt{Controller} process,
responsible for the migration between populations and keeping track of the algorithm's
iterations. The \texttt{Message Queue} process runs the \texttt{Input} and \texttt{Output}
queues. Finally, there is at least one \texttt{Stateless Function} process responsible
for running the isolated algorithms. In the example, two processes \texttt{PSO} and \texttt{GA},
are shown.

@@

\item Once they finish the execution, the population state
is again packed with additional metadata about the algorithm's execution.
The resulting populations are now pushed to the Output queue. Once finished, another 
population is pulled from the queue.

\item The \texttt{Controller} process is responsible for keeping track of the progress of 
the search. It pulls current populations from the \texttt{Output} queue, inspects the metadata. 
If an optimal solution has been found or the maximum number of iteration has been
reached, it signals the execution to stop.

@@

This reactive architecture has the following advantages:\begin{itemize}
\item 
An important aspect of the proposal is the decoupling of the population and 
the population-based algorithm. It is common that in a classic island-based algorithm,
each island is executed in a separate processing node, i.e., in a virtual machine, CPU,
or thread. In this case, we can have just one processing node or many nodes, each 
running with a different search strategy or parameters. 
\item 
Also, the reactive controller gives designers more control over the multi-population algorithm.
In this process, designers can dynamically change the number of populations, population parameters,
and migration details on-the-fly.
\item 
Another advantage is that algorithm designers have many options for implementing this simple architecture. 
The same components can be implemented as a single multi-threaded program or 
as a highly scalable serverless cloud application. Most modern languages include 
constructs for asynchronous programming using queues or channels for multi-threaded
execution.
\end{itemize}

@@

On the other hand, we have several caveats: It is more
costly to move entire populations as messages than passing only
specific individuals from a process to another. Designers must consider
this cost when working with large individuals, and if possible, send
populations using several messages, use compression, or adjust the populations'
size. Pool-based algorithms also suffer from this drawback.
%,but web-based implementations have been working with continuous
%optimization problems with 1000 dimensions.  However, in other cases, % Reference that paper - Mario
%this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four
multi-population-based algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
populations, processing nodes, and algorithms.  In Figure \ref{fig:population_message}, we show the
main components of a message-driven architecture. We can see that the algorithm
and populations are separated, and the \texttt{Population queue} keeps the state
of populations. While algorithms only need the populations as parameters
for their execution.

@@

In contrast, in the classic island model shown in Figure
\ref{fig:classicisland}, algorithms are  population methods, and the same processing
node keeps both the algorithm process and the state of the population.  Other
processing nodes follow the same configuration and execution form, and they
only pass specific individuals between them.  In this model, adding additional
processing nodes, on-runtime, can be more difficult because they need to have
an addressing mechanism or simply be aware of these new nodes. This is
solved in part in Peer to Peer (P2P) evolutionary algorithms
\cite{juanlu:europar}. However, coupling still exists, meaning that every
node must run the whole evolutionary algorithm and hold full copies of
its subpopulation.

@@

\begin{enumerate}
\item The controller continually pulls new messages from the \texttt{Output} 
message queue. These messages are instantly emitted by the  \texttt{consumed\_messages}
\texttt{Observable}.This particular example shows that the stream receives 
two populations from another experiment, and these are shown as two red marbles.  
The filter operator removes all messages that belong to a different experiment; 
in this case, we are only interested in blue marbles.

@@

Developers can add new algorithms as stateless functions in
several ways and with two levels of integration. In the first level of
integration, the algorithm execution is not dependent on an initial
configuration created in the setup, and it only needs to take messages from the
input queue and return the modified population to the output queue. At the
second level, the algorithm receives a configuration from the initial setup,
specifying the initial algorithm parameters. We show an example of the configuration
file received by workers in Listing \ref{code:config}  in which we abbreviated
the population key (the ellipsis is used instead). The message includes the 
population's current state and data about the problem, algorithm, and parameters.

@@

  \item[A container orchestration service]: These services are compatible with {\tt docker-compose}
    files for configuration and provide transparent provisioning. Examples of these
    services are Amazon Elastic Container Service,  Red Hat Openshift
    (which is also open-source), and Marantis Docker Enterprise using Docker Swarm.
    Salza et al. \cite{salza2019speed} tested a variant of this option by using a CoreOs cluster.

@@

%\subsubsection{Setup} % Even if it's a sub-sub section, flesh this out ... - JJ
%\label{setup:local-cloud}

We chose the second option, a container orchestration service. We
deployed the EvoSwarm application in Amazon Elastic Container Service (ECS)
since it needs fewer changes in configuration (from deploying it locally using
{\tt docker-compose}) and it has a lower cost than the Kubernetes
alternative.

We can define an ECS cluster directly on the AWS Elastic Cloud VMs or through the
AWS Fargate service. Fargate is a serverless compute engine for containers that
provides automatic provision and management of AWS ECS resources. With this
service, calculating the cost of execution is simplified because users only
specify the computing units and memory for the application, instead of every
component. However, AWS Fargate is not suited for an experimental setting because
many of the details are out of user control. For this reason, we chose the
alternative of using a cluster of EC2 containers. The drawback is that we need to manually
provision the  EC2 containers, networking, and group permissions for each
experimental setup; for instance, when going from 2 to 8 workers, we need to increase the
number of EC2 instances to meet the required amount of CPU units.
